Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Working on index and glossary

\documentclass{report}
\usepackage{fullpage}
\usepackage[nottoc,notlot,notlof]{tocbibind}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage{amsmath}

@@

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables and
synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically} because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}: instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

Model checking is not a replacement for formal verification.  Formal
verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be relatively
simple.  In particular, it needs to have a relatively small number of
reachable states.
If the model is too large, it may run for longer than we may care to
wait for or even run out of memory.

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  We call this a \emph{critical section}
(aka critical region): a
set of instructions where only one process is allowed to execute at a
time.

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.

@@

\chapter{Peterson's Algorithm}
\label{ch:peterson}

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
(the parentheses are mandatory) is a pointer to \texttt{x}
(also known as the \emph{address} of \texttt{x}).
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.

@@

\chapter{Semaphores}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections} (CCSs).
All critical sections are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}.
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \texttt{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
There are two operations on condition variables: \texttt{wait} and \texttt{signal}.
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

In the late 70s, Xerox PARC developed a new programming language called
Mesa~\cite{LR80}.
Mesa introduced various important concepts to programming languages,
including software exceptions and incremental compilation.  Mesa also
incorporated a version of monitors.
However, there are some subtle but important differences with Hoare
monitors that make Mesa monitors quite unlike split binary semaphores.

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variable that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation release the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
the baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point $q$ gets the baton,
assuming there even was such a process $q$.

@@

Mesa monitors provide another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Using \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
using \texttt{signal} on an empty bathroom.) Mesa also has a function called
\texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
a bedroom, let's all processes in that bedroom into the hall.

@@

\chapter{Actors}

@@

The actor synchronization model is popular in a variety of programming
languages, including Erlang and Scala.  Actor support is also available
through popular libraries such as Akka, which is available for various
programming languages.  In Python, Java, and C/C++,
actors can be easily emulated using threads and \emph{synchronized queues}
for messaging.  Each thread would have one such queue for receiving messages.
Dequeuing from an empty synchronized queue blocks the thread until
another thread enqueues a message on the queue.

@@

A \emph{macro step} can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.

@@

>>>>

\documentclass{report}
\usepackage{fullpage}
\usepackage{imakeidx}\makeindex
\usepackage{hyperref}
\usepackage[nottoc,notlot,notlof]{tocbibind}
\usepackage{glossaries}\makeglossaries
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage{amsmath}

@@

\newglossaryentry{macro step}
{
  name=macro step,
  description={is a programmable machine that receives input,
               stores and manipulates data, and provides
               output in a useful format}
}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables and
synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}
instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

Model checking is not a replacement for formal verification.
\index{formal verification}
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be relatively
simple.  In particular, it needs to have a relatively small number of
reachable states.
If the model is too large, it may run for longer than we may care to
wait for or even run out of memory.

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}: a
set of instructions where only one process is allowed to execute at a
time.

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) is called \emph{thread-safe}.

@@

\chapter{Peterson's Algorithm}
\label{ch:peterson}
\index{Peterson's Algorithm}

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
(the parentheses are mandatory) is a pointer to \texttt{x}
(also known as the \emph{address} of \texttt{x}).
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.

@@

\chapter{Semaphores}
\index{semaphore}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections} (CCSs)~\cite{Hoare73}.
All critical sections are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \texttt{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
\index{condition variable}
There are two operations on condition variables: \texttt{wait}
\index{wait}
and
\texttt{signal}.
\index{signal}
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

In the late 70s, Xerox PARC developed a new programming language called
Mesa~\cite{LR80}.
\index{Mesa}
Mesa introduced various important concepts to programming languages,
including software exceptions and incremental compilation.  Mesa also
incorporated a version of monitors.
However, there are some subtle but important differences with Hoare
monitors that make Mesa monitors quite unlike split binary semaphores.

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variable that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation release the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
the baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point $q$ gets the baton,
assuming there even was such a process $q$.

@@

Mesa monitors provide another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Using \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
using \texttt{signal} on an empty bathroom.) Mesa also has a function called
\texttt{notifyAll}
\index{notifyAll}
(aka \texttt{broadcast)})
\index{broadcast}
that, when applied to
a bedroom, let's all processes in that bedroom into the hall.

@@

\chapter{Actors and Synchronized Queues}
\index{actor}

@@

The actor synchronization model is popular in a variety of programming
languages, including Erlang and Scala.  Actor support is also available
through popular libraries such as Akka, which is available for various
programming languages.  In Python, Java, and C/C++,
actors can be easily emulated using threads and \emph{synchronized queues}
(aka \emph{blocking queues}) for messaging.
Each thread would have one such queue for receiving messages.
Dequeuing from an empty synchronized queue blocks the thread until
another thread enqueues a message on the queue.

@@

A \emph{macro step}
\glsadd{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.

@@

\printindex

% \glsaddall
\printglossaries
